---
title: "Theoretical background of the mhn package"
output:
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Theoretical background of the mhn package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 72
)
set.seed(1)
```

```{r library, include = FALSE}
library(mhn)
```

## Scope

This vignette is the theoretical companion to *Introduction to the mhn
Package*. The *Introduction* shows what the four exported functions
(`dmhn` / `pmhn` / `qmhn` / `rmhn`) and their moment helpers do; this
note develops the mathematical and algorithmic theory the package
implements. We summarise the Modified Half-Normal (MHN) family, the
Fox–Wright $\Psi$ function that appears in its normalising constant,
and the two sampling schemes the package draws on — Sun, Kong & Pal
(2023) and Gao & Wang (2025).

We deliberately keep the treatment compact. Full proofs and the
finer numerical considerations live in the original papers (see
References). Two implementation-relevant points — the typesetting
errata in Sun et al. (2023) that the package silently corrects,
and the benchmark-driven rationale for the
`rmhn(method = "auto")` crossover at 25 — are inserted in §4 and
§7 where they are most relevant.

## The MHN($\alpha$, $\beta$, $\gamma$) family

The MHN distribution has support $(0, \infty)$ and density

$$f(x \mid \alpha, \beta, \gamma) =
  \frac{2 \beta^{\alpha/2} \, x^{\alpha-1} \, \exp(-\beta x^2 + \gamma x)}
       {\Psi[\alpha/2,\, \gamma/\sqrt{\beta}]},
  \qquad x > 0,$$

with $\alpha > 0$, $\beta > 0$, $\gamma \in \mathbb{R}$.
The three parameters control three independent factors of the
density kernel:

| Parameter   | Factor it controls           | Role          |
|-------------|------------------------------|---------------|
| $\alpha$ | $x^{\alpha-1}$          | shape         |
| $\beta$  | $\exp(-\beta x^2)$      | Gaussian tail |
| $\gamma$ | $\exp(\gamma x)$        | exponential tilt |

Three familiar distributions sit inside the family as exact
special cases (Sun et al. 2023, Lemma 6):

| Constraint                 | Reduction |
|------|----------------------|
| $\gamma = 0$               | $X^2 \sim \mathrm{Gamma}(\alpha/2,\, \beta)$, i.e.\ sqrt-Gamma |
| $\alpha = 1$               | Truncated normal $\mathrm{TN}(\gamma/(2\beta),\, 1/\sqrt{2\beta},\, 0,\, \infty)$ |
| $\alpha = 1,\, \gamma = 0$ | Half-normal $\lvert Z \rvert$ with $Z \sim N(0,\, 1/(2\beta))$ |

There is also a degenerate *boundary* limit, not listed in the
table: as $\beta \to 0^+$ with $\gamma < 0$, the MHN density
converges to $\mathrm{Gamma}(\alpha,\, -\gamma)$. The package does
not handle this limit specially because $\beta > 0$ is required by
the MHN parameter space; for a Gamma draw with $\beta = 0$, use
`stats::rgamma` directly.

For each of the three *finite* special cases in the table above —
$\gamma = 0$, $\alpha = 1$, and both together — the package
detects the constraint within `sqrt(.Machine$double.eps)` tolerance
and dispatches to the corresponding closed-form primitive in
`stats`, so all four exported functions
(`dmhn` / `pmhn` / `qmhn` / `rmhn`) are exact in those three
regimes, bypassing the general series-and-rejection machinery.

```{r density-shape, echo = -c(2, 8), fig.cap = "MHN density as alpha crosses the log-concavity threshold; beta = 1, gamma = 1 fixed."}
x  <- seq(0.001, 4, length.out = 400)
op <- par(mar = c(4, 4, 2, 1))
plot(x, dmhn(x, alpha = 2,   beta = 1, gamma = 1), type = "l", lwd = 2,
     ylim = c(0, 1.4), ylab = "f(x)",
     main = "MHN density at beta = 1, gamma = 1")
lines(x, dmhn(x, alpha = 5,   beta = 1, gamma = 1), lwd = 2, col = "steelblue")
lines(x, dmhn(x, alpha = 1,   beta = 1, gamma = 1), lwd = 2, col = "tomato")
lines(x, dmhn(x, alpha = 0.5, beta = 1, gamma = 1), lwd = 2, col = "darkorange")
legend("topright", bty = "n", lwd = 2,
       col = c("darkorange", "tomato", "black", "steelblue"),
       legend = c("alpha = 0.5  (no interior mode)",
                  "alpha = 1    (truncated normal)",
                  "alpha = 2    (log-concave)",
                  "alpha = 5    (log-concave, shifted right)"))
par(op)
```

The figure illustrates the qualitative claims developed in the
next section. The density is log-concave whenever $\alpha \geq 1$
(Sun et al. 2023, Lemma 3a). For $\alpha > 1$ it has a finite
interior mode in closed form (Sun et al. 2023, Lemma 3b); the
borderline case $\alpha = 1$
reduces to the truncated normal, whose mode is
$\max(0,\, \gamma/(2\beta))$ — interior when $\gamma > 0$, on the
boundary $x = 0$ otherwise. For $\alpha < 1$ the density is
typically monotone decreasing with a boundary singularity at
$x \to 0^+$. The one exception is $\gamma > 0$ together with
$\alpha \geq 1 - \gamma^2/(8\beta)$, where the density is not
unimodal: it diverges at $x \to 0^+$, dips to a local minimum,
rises to a single interior local maximum, and then decays
(Sun et al. 2023, Lemma 3c). At $(\beta, \gamma) = (1, 1)$ this
threshold
is $1 - 1/8 = 0.875$, so the orange ($\alpha = 0.5$) curve in
the figure sits in the strict monotone-decreasing regime, with
no interior local maximum.

## Shape and moments

Three facts from Sun et al. (2023) drive everything the package
does with shape, moments, and the mode.

**Log-concavity is a function of $\alpha$ alone** (Sun et al. 2023,
Lemma 3a). For $\alpha \geq 1$ the density is log-concave
irrespective of
$\beta$ and $\gamma$; for $\alpha < 1$ it is not. This
single threshold is what splits the Gao & Wang (2025) RTDR sampler
into its log-axis and original-axis branches in §6 below.

**The mode has a closed form for $\alpha > 1$** (Sun et al. 2023,
Lemma 3b):

$$X_{\mathrm{mode}}
 = \frac{\gamma + \sqrt{\gamma^2 + 8\beta(\alpha - 1)}}{4\beta}.$$

For $\alpha = 1$ the mode is $\max(0,\, \gamma/(2\beta))$
(the truncated-normal mode, Sun et al. 2023, Lemma 6b). For
$\alpha < 1$
the density is either monotone decreasing or develops a non-trivial
local maximum and local minimum, depending on the sign of
$1 - \gamma^2/(8\beta) - \alpha$ (Sun et al. 2023, Lemma 3c–d).
The helper
`mhn_mode()` implements this case split and returns `NA` when no
interior mode exists.

**Moments satisfy a two-term recurrence** (Sun et al. 2023,
Lemma 2b):

$$E(X^{k+2})
 = \frac{\alpha + k}{2 \beta}\, E(X^{k})
 + \frac{\gamma}{2 \beta}\, E(X^{k+1}).$$

Together with the closed-form ratio for $E(X)$ in terms of
$\Psi$ (Sun et al. 2023, Lemma 2a), this gives every higher
moment.
The helpers `mhn_mean()`, `mhn_var()`, `mhn_skewness()` and
`mhn_kurtosis()` apply the recurrence in C++ via Rcpp; the *Introduction*
vignette shows them in action. A useful sanity bound
(Sun et al. 2023, Lemma 4c) is $\mathrm{Var}(X) \leq 1/(2\beta)$
for every
$\alpha \geq 1$, $\gamma \in \mathbb{R}$.

## The Fox–Wright $\Psi$ normalising constant

The normalising constant in the MHN density is

$$\Psi\!\left[\frac{\alpha}{2},\, z\right]
 \;=\; \sum_{n=0}^{\infty}
        \frac{\Gamma\!\big(\tfrac{\alpha + n}{2}\big)}{n!} \, z^{n},
 \qquad z = \frac{\gamma}{\sqrt{\beta}}.$$

This series is absolutely convergent for every
$\alpha > 0$ and every $z \in \mathbb{R}$, and is strictly
positive. It depends on $(\alpha, \beta, \gamma)$ only through
the two-dimensional combination $(\alpha, z)$.

A useful organisational point: **$\Psi$ matters only for some
of the package's exported functions**.

| Function group                            | Needs $\Psi$? | Why |
|-------------------------------------------|:-----------------:|-----|
| `dmhn`, `pmhn`, `qmhn`                    | yes               | $\Psi$ sits in the density's denominator and in the CDF series. |
| `mhn_mean`, `mhn_var`, `mhn_skewness`, `mhn_kurtosis` | yes  | Moments are ratios $\Psi[(\alpha+k)/2,\, z] / \Psi[\alpha/2,\, z]$. |
| `rmhn` (Sun or RTDR path)                 | no                | $\Psi$ enters both proposal scale and target as a common factor that cancels. |

The practical consequence is that any numerical issues in evaluating
$\Psi$ — accumulated rounding in `lgamma`, or truncation error
in the series — can only affect `d/p/q` and the moment helpers; they
cannot leak into the sampler.

For evaluation, the package follows the case split of Sun et al.
(2023, Lemma 9–11):

- $\gamma = 0$ gives $\Psi[\alpha/2, 0] = \Gamma(\alpha/2)$.
- $\alpha = 1$, and the $\alpha = 2,\, \gamma \geq 0$ corner,
  both reduce to closed forms involving $\Phi$ (Sun et al. 2023,
  Lemma 9c).
- $\gamma > 0$ uses the alternating-free series of Sun et al.
  (2023, Lemma 10) with a pre-computed truncation point. All terms
  are positive, so there is no catastrophic cancellation.
- $\gamma < 0$ switches to the integral representation
  (Sun et al. 2023, Lemma 11) and uses Boost's `gauss_kronrod`
  (for
  $\alpha \geq 1$) or `tanh_sinh` (for $\alpha < 1$, where the
  boundary singularity $u^{\alpha-1}$ needs the double-exponential
  rule).

### Errata in Sun et al. (2023)

The published paper has two typesetting errors that touch the
implementation. Each is corrected in the supplementary derivation,
and the package follows the corrected form.

- **Mixture weights** $p_{i}$ **in Lemma 7.** The main-text
  statement of Lemma 7 gives the weights for the
  $\sqrt{\mathrm{Gamma}}$ mixture representation as
  $p_{i} = \tfrac{1}{2\beta^{\alpha/2}} \cdot
   \tfrac{\Gamma((\alpha + 1 + i)/2)\, (\gamma/\sqrt{\beta})^{i}}
        {i!\, \Psi[\alpha/2,\, \gamma/\sqrt{\beta}]}$.
  These do not sum to one, contradicting the definition of $\Psi$
  and the Poisson–Gamma hierarchical representation in
  Sun et al. (2023, Lemma 5a). The correct form, reproduced in the
  supplementary
  proof of Lemma 7 (Supplementary §2.13), is
  $$p_{i} \;=\;
   \frac{\Gamma((\alpha + i)/2)\, (\gamma/\sqrt{\beta})^{i}}
        {i!\, \Psi[\alpha/2,\, \gamma/\sqrt{\beta}]}$$
  i.e.\ drop the leading $1/(2\beta^{\alpha/2})$ and use
  $\alpha + i$ in the $\Gamma$ argument instead of
  $\alpha + 1 + i$. This affects Algorithm 2 only, which the
  package does not implement; the correction is recorded here for
  completeness.

- **Series truncation discriminants** $C_{1}$, $C_{2}$ **in
  Lemma 10.** The truncation point of the $\Psi$ series is the
  smallest $k$ for which $A(k+1)/A(k) \leq q$ (with an analogous
  bound for $B$). Reducing this to a quadratic in $k$ gives a
  discriminant whose last term is $\alpha x^{2}$ (and
  $(\alpha+1) x^{2}$ for $C_{2}$). The main-text statement of
  Lemma 10 prints these as $\alpha x$ and $(\alpha+1) x$ — a
  single missing power of $x$. The supplementary derivation
  (Supplementary §2.2.1, §2.2.2) carries $z^{2}$ throughout, in
  agreement with the rederivation. The corrected $x^{2}$ form is
  what the package uses in
  [src/mhn_psi_series.cpp](https://github.com/t-momozaki/mhn/blob/main/mhn/src/mhn_psi_series.cpp)
  (marked with an `// ERRATA` comment); evaluating $\Psi$ at the
  published $x$ form would give a too-small truncation point and
  silently lose precision.

## Sampling: Sun et al. (2023)

Sun et al. (2023) propose three Accept–Reject schemes, indexed by
the sign of $\gamma$ and a coarse split on $\alpha$. Two of
them — Algorithm 1 and Algorithm 3 — are implemented in the package
and reachable via `rmhn(method = "sun")`. Algorithm 2 is not, and
this section explains the reason.

### Algorithm 1 — $\gamma > 0$, $\alpha \geq 1$

For each parameter triple, Algorithm 1 constructs **two** proposal
distributions — a Normal with mean
$\mu_{\mathrm{opt}}$ and variance $1/(2\beta)$, and a
$\sqrt{\mathrm{Gamma}(\alpha/2,\, \delta_{\mathrm{opt}})}$ —
each with a multiplicative envelope constant
$K_{1}$ and $K_{2}$. Setup chooses whichever envelope is
tighter ($K_{1} \leq K_{2}$ or the reverse). The two optima
$\mu_{\mathrm{opt}}$ and $\delta_{\mathrm{opt}}$ have closed
forms (Sun et al. 2023, Theorem 1b), so setup is cheap. The
acceptance probability is bounded below by 0.8 for $\alpha \geq 4$
(Sun et al. 2023, Theorem 2e). For $1 \leq \alpha < 4$ no uniform
lower
bound is proved, but the original Figure 2 shows empirically high
acceptance throughout that range.

### Algorithm 3 — $\gamma \leq 0$

When the tilt is non-positive, Theorem 3 of Sun et al. (2023)
supplies a proposal kernel based on an AM–GM-type inequality. With a tuning point
$m > 0$ and the exponent
$r = (\beta m + |\gamma|)/(2\beta m + |\gamma|)$, the candidate is

$$T \sim \mathrm{Gamma}\!\big(\alpha \cdot r,\, m(\beta m + |\gamma|)\big),
\qquad X = m \cdot T^{r}.$$

The construction works for every $\alpha > 0$; the uniform
acceptance bound $\geq 1/\sqrt{2} \approx 0.707$ (Sun et al. 2023,
Theorem 4c) is proved only for $\alpha > 1$, but the procedure
also samples correctly when $\alpha \leq 1$. The matching point is
initialised from a closed-form heuristic of Sun et al. (2023, §4.2):
$m_{\mathrm{init}} = \alpha^{2}/(1 + \alpha)$ for $\alpha \leq 1.1$
and a more elaborate expression based on the mode and the rightmost
inflection point for $\alpha > 1.1$. The package then applies a
single Newton-Raphson refinement step in both regimes, producing
$m_{\mathrm{recommend}}$, whose acceptance probability is within
$0.2\%$ of the optimum across the parameter range tabulated in
Sun et al. (2023, Table 1).

### Why Algorithm 2 is not implemented

Algorithm 2 targets $\gamma > 0$ with $\alpha < 1$ via the
mixture representation

$$f_{\mathrm{MHN}}(x \mid \alpha, \beta, \gamma)
 \;=\; \sum_{i=0}^{\infty} p_{i}\, f_{i}(x \mid \alpha, \beta),
\qquad
f_{i} \;\sim\; \sqrt{\mathrm{Gamma}\!\big((\alpha + i)/2,\, \beta\big)}.$$

The mixture-component sampler needs a truncation point $M^{\dagger}$
chosen so that the tail mass beyond $M^{\dagger}$ is below a
target tolerance (Sun et al. 2023, Lemma 8). For large
$\gamma^{2}/\beta$
this truncation grows like $\gamma^{2}/(\varepsilon^{2} \beta)$,
and the whole scheme degrades catastrophically — Gao & Wang (2025,
Table 1) report a slowdown of up to $\sim 3 \times 10^{4}\times$
against RTDR in the most extreme case
$(\alpha, \gamma) = (0.01, 10000)$. The package therefore declines to implement
Algorithm 2 and routes $(\alpha < 1, \gamma > 0)$ to RTDR, where
the acceptance bound $\geq 1/e$ (next section) holds uniformly.

## Sampling: Gao & Wang (2025) RTDR

Gao & Wang (2025) introduce a *Relaxed* Transformed Density
Rejection (RTDR) sampler that comes with a uniform lower bound on
the acceptance probability across the *entire* parameter space.
Three ideas drive the construction.

### $T_{c}$-concavity

Standard TDR builds piecewise-linear envelopes on top of a
log-transformed density. RTDR generalises to the $T_{c}$ family

$$T_{c}[x] = \mathrm{sign}(c) \cdot x^{c} \quad (c > -1,\, c \neq 0),
\qquad T_{0}[x] = \log x.$$

A density $f$ is *$T_{c}$-concave* if $T_{c}[f]$ is concave.
Logarithmic concavity (the usual TDR setting) is the
$c = 0$ special case. The package's RTDR implementation uses
$c = -1/2$ on the log-axis density for the $\alpha < 1$
branches because $T_{-1/2}[x] = -x^{-1/2}$ accommodates the
slower decay of $g(y)$ when the original $f$ has a boundary
singularity at $x = 0$.

### Working on the log axis when $\alpha < 1$

When $\alpha < 1$ the density $f(x \mid \alpha, \gamma)$ on
$(0, \infty)$ blows up at the left endpoint and is not amenable
to direct envelope construction. The change of variable
$y = \log x$ yields

$$g(y \mid \alpha, \gamma)
 \;=\; \exp\!\big( \alpha y - e^{2y} + \gamma e^{y} \big),
\qquad y \in \mathbb{R},$$

which is bounded and unimodal (Gao & Wang 2025, Lemma 3.3). RTDR
operates on $g$ for $\alpha < 1$ and on $f$ directly for
$\alpha \geq 1$.

### The four regions

After scale normalisation $\sqrt{\beta} X \to X$
(so we may assume $\beta = 1$), the RTDR envelope construction
splits into four regions based on $(\alpha, \gamma)$. The $\gamma$
in the table below is the normalised tilt
$\gamma_{\mathrm{norm}} = \gamma/\sqrt{\beta}$; for a general
$\beta$ the package's `rmhn(method = "rtdr")` rescales internally
and compares $\gamma_{\mathrm{norm}}$ against the boundary
$2(1 - \sqrt{1 - 2\alpha})$.

| Region | $\alpha$        | $\gamma$             | Envelope target              | Reference                                          |
|:------:|-----------------|----------------------|------------------------------|----------------------------------------------------|
| (a)    | $\geq 1$        | any                  | $f$ log-concave              | Gao & Wang (2025, Theorem 3.1)                     |
| (b)    | $[1/2,\, 1)$    | any                  | $g$ $T_{-1/2}$-concave       | Gao & Wang (2025, Corollary 3.1; Theorem 3.2)      |
| (c)    | $(0,\, 1/2)$    | $\leq 2(1 - \sqrt{1 - 2\alpha})$ | $g$ $T_{-1/2}$-concave | Gao & Wang (2025, Lemma 3.5; Theorem 3.2)        |
| (d)    | $(0,\, 1/2)$    | $> 2(1 - \sqrt{1 - 2\alpha})$    | $g$ with inflection at $y^{*} = \log(\gamma/4)$ | Gao & Wang (2025, Theorem 4.4)      |

Region (a) is the easiest: $f$ is log-concave, two tangent lines
and a constant segment suffice. Regions (b) and (c) reuse the same
$T_{-1/2}$ envelope on $g$, with only the validity argument
differing. Region (d) is the hardest — $g$ has an inflection
point and the envelope needs $K$ secant segments on the
log-convex side, with $K$ growing logarithmically in $\gamma$.

### Why "relaxed"

The classical TDR construction asks for *optimal* tangent contact
points, which generally have no closed form and must be found by
Newton iteration. Gao & Wang (2025, Theorem 2.1) show that the
acceptance bound $\leq e \approx 2.719$ (equivalently,
acceptance probability $\geq 1/e \approx 0.368$) holds as long as
the contact points lie *anywhere* inside a fixed interval — namely

$$\log\!\frac{f(m)}{f(t)} \in
 \begin{cases}
   [0.46,\, 2.49] & f \text{ log-concave (region (a))}, \\
   [0.93,\, 1.99] & g \text{ $T_{-1/2}$-concave (regions (b), (c))}.
 \end{cases}$$

Setup is therefore a handful of cheap iterations that terminate as
soon as the bracket is hit, with no need for a precise Newton
solution. This is the main practical edge of RTDR over
Algorithms 1 / 3 of Sun et al. (2023) in Gibbs-style workloads
where
$(\alpha, \beta, \gamma)$ changes every iteration: the setup cost
is amortised over a single draw instead of many, and a cheap setup
wins.

Composing the bounds in Gao & Wang (2025, Theorems 3.1, 3.2,
4.4) yields the
headline guarantee of RTDR: for **every**
$(\alpha, \beta, \gamma)$ with $\alpha > 0$,
$\beta > 0$, $\gamma \in \mathbb{R}$, the acceptance
probability is at least $1/e \approx 0.368$. The three
algorithms of Sun et al. (2023) only have parameter-dependent
guarantees (and only on parts of the parameter space — see §5).

A quick sanity check that the RTDR sampler converges to the right
distribution in all four regions — empirical means from RTDR vs
theoretical means across regions (a), (b), (c), (d), each based on
5000 draws:

```{r rtdr-regions}
cases <- data.frame(
  region = c("(a)", "(b)", "(c)", "(d)"),
  alpha  = c(2.0,    0.7,    0.3,    0.3),
  gamma  = c(0.5,    1.0,    0.5,    5.0)
)
set.seed(42)
empirical <- mapply(
  function(a, g) mean(rmhn(5000, alpha = a, beta = 1, gamma = g, method = "rtdr")),
  cases$alpha, cases$gamma
)
theoretical <- mapply(mhn_mean, cases$alpha, beta = 1, cases$gamma)
data.frame(cases,
           empirical_mean   = round(empirical,   4),
           theoretical_mean = round(theoretical, 4))
```

The four rows hit the right means to two-to-three decimal places —
about the sampling tolerance at $n = 5000$.

## The `rmhn(method = "auto")` dispatch

`rmhn` exposes three values of `method`: `"sun"` (force the Sun
Algorithm 1 / 3 path), `"rtdr"` (force the Gao & Wang RTDR path),
and `"auto"` (default). The `"auto"` route is the package's
recommendation; it composes the four shortcuts above into one
decision tree (this is the Step 3.7.3 final form, confirmed by the
benchmarks under `inst/benchmarks/`):

```
                  rmhn(n, alpha, beta, gamma, method = "auto")
                                    │
                                    ▼
       ┌─ alpha = 1, gamma = 0  →  half-normal       (Sun et al. 2023, Lemma 6c)
       │
       ├─ gamma = 0             →  sqrt-Gamma        (Sun et al. 2023, Lemma 6a)
       │
       ├─ alpha = 1             →  truncated normal  (Sun et al. 2023, Lemma 6b)
       │
       ├─ alpha < 1 and gamma > 0
       │      →  RTDR  (regions (b) / (c) / (d), Gao & Wang 2025, Theorems 3.2, 4.4)
       │
       ├─ alpha > 1 and gamma > 0
       │      →  Sun Algorithm 1                    (Sun et al. 2023, Theorems 1, 2)
       │
       └─ gamma < 0
              ├─ samples_per_setup ≥ 25  →  RTDR    (region (a))
              └─ samples_per_setup < 25  →  Sun Algorithm 3
```

Write $S$ for the `samples_per_setup` quantity in the last branch,

$$S \;=\; \max\!\Big(1,\; \frac{n}{\max(L_{\alpha},\, L_{\beta},\, L_{\gamma})}\Big),$$

where $L_{\alpha}$, $L_{\beta}$, $L_{\gamma}$ are the
recycled lengths of the three parameter vectors. The intuition
follows from decomposing the expected time per draw as

$$E[\text{time}/\text{draw}]
 \;=\; \frac{T_{\text{setup}}}{n}
       \;+\; C_{\text{reject}} \times T_{\text{per-proposal}},$$

where $T_{\text{setup}}$ is the one-off cost of preparing the
proposal (a single $m_{\text{init}}$ formula plus at most one
Newton step for Sun Algorithm 3; an iterative contact-point search
for RTDR), $C_{\text{reject}} = 1/\text{acceptance}$ is the
expected number of proposals per accepted draw, and
$T_{\text{per-proposal}}$ is the cost of one proposal cycle (sample
plus acceptance log-ratio evaluation).

For $\gamma < 0$ the two samplers sit on opposite ends of this
trade-off:

- **Sun Algorithm 3** has a *cheap setup* — $m_{\text{init}}$ has
  a closed form, and the optional Newton refinement uses one
  `boost::math::digamma` evaluation — but a *more expensive
  per-proposal*: every proposal calls `rgamma`, evaluates the
  fractional power $X = m \cdot T^{r}$, and the acceptance
  log-ratio carries the same $(X/m)^{1/r}$ term.
- **RTDR (region (a))** has a *more expensive setup*: each
  contact point $t_{l}$, $t_{r}$ is found by an iterative search
  that terminates as soon as $\log f(m)/f(t)$ lands in the
  bracket $[0.46,\, 2.49]$. But the *per-proposal* is light:
  sampling from the piecewise envelope is straight arithmetic plus
  a single `log`, and the acceptance ratio adds only two more
  `log`s.

When $S$ is large the per-proposal term dominates and RTDR wins;
when it is small — the Gibbs case where each iteration redraws
$(\alpha, \beta, \gamma)$ — the setup term dominates and Sun
Algorithm 3 wins. The crossover at 25 is
the round value chosen inside the empirical break-even window
$n \in [10,\, 25]$ observed at every one of the 20 surveyed
$(\alpha, \gamma)$ cells in
[inst/benchmarks/auto_dispatch.R](https://github.com/t-momozaki/mhn/blob/main/mhn/inst/benchmarks/auto_dispatch.R).

A small concrete demonstration: with `set.seed` aligned,
`method = "auto"` and the *expected* forced method produce
bit-identical draws on each branch.

```{r auto-dispatch}
# Branch: alpha < 1 and gamma > 0  -->  expects RTDR
set.seed(7); auto1 <- rmhn(20, alpha = 0.7, beta = 1, gamma = 2)
set.seed(7); rtdr1 <- rmhn(20, alpha = 0.7, beta = 1, gamma = 2, method = "rtdr")
identical(auto1, rtdr1)

# Branch: alpha > 1 and gamma > 0  -->  expects Sun Algorithm 1
set.seed(7); auto2 <- rmhn(20, alpha = 3,   beta = 1, gamma = 2)
set.seed(7); sun2  <- rmhn(20, alpha = 3,   beta = 1, gamma = 2, method = "sun")
identical(auto2, sun2)

# Branch: gamma < 0 with samples_per_setup < 25  -->  expects Sun Algorithm 3
set.seed(7); auto3 <- rmhn(5,  alpha = 2,   beta = 1, gamma = -1)
set.seed(7); sun3  <- rmhn(5,  alpha = 2,   beta = 1, gamma = -1, method = "sun")
identical(auto3, sun3)
```

The three `TRUE` outputs are the same invariant that
`tests/testthat/test-rmhn.R` enforces as a regression test.

## References

Sun, J., Kong, M., & Pal, S. (2023). The Modified-Half-Normal
distribution: Properties and an efficient sampling scheme.
*Communications in Statistics -- Theory and Methods*, 52(5),
1507--1536. \doi{10.1080/03610926.2021.1934700}

Gao, F. & Wang, H.-B. (2025). Generating modified-half-normal random
variates by a relaxed transformed density rejection method.
*Communications in Statistics -- Simulation and Computation*.
\doi{10.1080/03610918.2025.2524551}

Robert, C. P. (1995). Simulation of truncated normal variables.
*Statistics and Computing*, 5(2), 121--125.
